---
title: "Lab 11"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

```{r}
library(tidyverse)
```

In this lab we re-do some of lab 10
using the `map()` functions from `purrr`.
Recall that the folder `Lab10Data contains several CSV data files.

```{r}
dfiles <- dir("Lab10Data",full.names=TRUE)
dfiles
```

1. Write a function `read_rename_csv()` 
that (i) reads in a CSV file with `read_csv()`
and (ii) changes the names of the columns
of the resulting tibble to `c("x","y")`.
(Compare to question 2 of lab 10.)


2. Use `map()` and `read_rename_csv()`
to read and rename all 9 
files from the Lab10Data folder.


3. Re-do your call to `map()` from part 2.
This time, define the function that reads 
and renames the data files on the fly, 
using  `~` and `.`. 
Do you prefer the approach of question 2 or
of question 3 (this question)?

Hint: Use 
`~{ ... }` to define your function, where
`...` denotes the body of the function you will
define. Remember that `x;y` is equivalent 
to having `x` and `y` on two lines of R code.

4. We will now apply the forward pipe several times
to get an equivalent to the `plot.study_data()` function 
you wrote in lab 10. The steps to take are
    i. Pipe `dfiles` through a call to  `map()` that 
    reads and renames the files (use your code from either
    question 2 or 3, whichever you preferred),
    ii. pipe the result through `bind_rows(.id="study")`
    (read the documentation for `bind_rows()`),
    iii. pipe the result through `mutate()` to 
    change `study` to a factor, and 
    iv. pipe the result into `ggplot()` to make the plot.
    