---
title: "Lab 10"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

The folder `Lab10Data contains several CSV data files.

```{r}
dfiles <- dir("Lab10Data",full.names=TRUE)
dfiles
```

1. Write R code to read in the first file.
Print the tibble that you just read in.
Use `names()` to change the column names of the
tibble to `x` and `y`.
Repeat for the second file. How many observations
are in these first two files?

2. Use `vector()` to create an empty vector 
called `ff` that is of mode "list" and length 9.
Now write a `for()` loop to loop over the 9 files in
`dfile` and for each (i) read the file in to a 
tibble, and change the column names to 
`x` and `y` as in part (1), and  (ii) copy the 
tibble to an element of your list `ff`.

3. Write a function called `read.study_data`
that takes a vector of data file
names (like `dfile`) as input, reads the data files
into a list, assigns class "study_data" to the list,
and returns the list. Your function should use `length(dfile)`
to determine the number of files.

4. Write a function `plot.study_data()` that 
takes an object of class "study_data" as input.
The first 5 lines of your function should be the
following, which creates a tibble  with
columns `study`, `x` and `y`:
```{r,eval=FALSE}
  dat <- NULL
  for(i in seq_along(ff)) {
    d <- ff[[i]]
    dat <- rbind(dat,tibble(study=i,x=d$x,y=d$y))
  }
```
Have your function coerce `study` to a factor, and then call
`ggplot()` to make a plot of `y` *versus* `x`,
with different colours for the different studies.
Add points and smoothers to your plot.

