---
title: "Stat 261, lab 2"
author: "Brad McNeney"
output: 
  html_document:
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## HIV prevalence from WHO

- Estimated HIV prevalence was obtained from the 
`gapminder` website https://www.gapminder.org/data/
    * Estimated number of people living with HIV per 100 population of age group 15-49.
    * Original data source is the UNAIDS online database at 
    http://www.aidsinfoonline.org
    
- A spreadsheet of the data are available in the Labs folder of the class GitHub repository. We 
can read in these data as follows (we'll learn about reading in data in week 5):

```{r}
library(tidyverse)
hiv <- read.csv("HIVprev.csv",stringsAsFactors = FALSE)
hiv <- select(hiv,Country, year, prevalence)
```

Take a look at the top and bottom few
lines of raw data.

```{r}
head(hiv)
tail(hiv)
summary(hiv)
```

1. Plot time series of HIV prevalence by year for each country.


2. Redo the above plot but experiment with different `alpha` values. What problem
does setting a small `alpha` overcome? What feature of the graph 
is hidden when we do not set `alpha`?

3. In the following code chunk we create a new dataset comprised of countries that had HIV prevalence 
greater than 10% in one or more of the years monitored (we will learn about this kind of 
"data wrangling" in week 5).

```{r}
cc <- c("Botswana","Central African Republic","Congo","Kenya","Lesotho","Malawi","Namibia","South Africa","Swaziland","Uganda","Zambia","Zimbabwe")
hihiv <- filter(hiv,Country %in% cc)
```

Add red lines for the above countries to your time series plot.
