---
title: "Stat 261, lab 3"
author: "Brad McNeney"
output: 
  html_document:
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## HIV prevalence from WHO

- The HIV prevalence data from lab 2 was modified from its raw form. 
In this lab we will work with the raw data.

- A spreadsheet of the data `HIVprevRaw.csv` is available in the Labs folder of the class GitHub repository. We 
can read in these data as follows (we'll learn about reading in data in week 5):

```{r}
library(tidyverse)
hiv <- read.csv("HIVprevRaw.csv",stringsAsFactors = FALSE)
```

- Take a look at the top few lines of raw data.

```{r}
head(hiv)
```

- Make a copy of `hiv` for use in exercise 4.

```{r}
hivcopy <- hiv
```


- In exercises 1 - 3, save the results of each data manipulation in 
`hiv`. In exercise 4  you will use the copy `hivcopy`.

1. The first column of the data frame is the country, but it
has been named`Estimated.HIV.Prevalence.....Ages.15.49.1`.
Use the `rename()` function to rename this column  `Country`.


2. The data from 1979 to 1989 is very sparse. Remove these
columns from the data frame.


3. Sort the data in descending order of prevalence in 2011. Print the first 6 rows of 
your final data set.


4. Use the copy `hivcopy` and the forward pipe to chain the data manipulations 
of exercises 1-3.
