---
title: "Lab 5"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

```{r}
library(tidyverse)
```


## HIV prevalence from WHO

* We used a tidy version of the HIV prevalence data in lab 2, and
saw the raw version in lab 3. In this lab we will tidy the latter into the former.

```{r,message=TRUE}
hiv <- read_csv("HIVprevRaw.csv")
```

(The columns for 1988 and 1989 are completely empty and 
were read in as logical. We will be removing these and so won't
worry about over-riding the logical with double.)


1. The first column of the data frame is the country, but it
has been named`Estimated HIVPrevalence% - (Ages 15 49 1)`.
Use the `rename()` function to rename this column  `Country`.
(Hint: The current variable name contains special characters
and will need to be enclosed in quotes.)


2. The data from 1979 to 1989 is very sparse. Remove these
columns from the data frame.

3. Gather the yearly prevalence estimates into key, value pairs
with `year` as the key and `prevalence` as the value. When
you gather, remove explicity missing values. After gathering,
sort on "Country". 

