---
title: 'Stat 260, Lecture 3, Data Visualization II'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
data(diamonds)
library(dplyr)
library(gapminder)
data(gapminder)
gapminder <- mutate(gapminder, 
                    log10Pop = log10(pop), 
                    log10GdpPercap = log10(gdpPercap))
```

## References

- Chapter 1 of printed text, Chapter 3 of online text.
- ggplot2 cheatsheet at [https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf]
- Wickham (2009) ggplot2: Elegant graphics for data analysis, Chapters 4 and 5.
- Chang (2012) R graphics cookbook. Available at  [http://www.cookbook-r.com/Graphs/]

## More `ggplot` Examples

Best way to learn data visualization is look at examples, such as those in the references on the 
previous slide. In this lecture, more examples to illustrate working with `ggplot` layers:
- data, 
- aesthetic mapping, 
- geom, 
- statistical transformation and
- position adjustment

## Diamonds dataset

\scriptsize

```{r}
p <- ggplot(diamonds,aes(x=carat,y=price,colour=cut)) +
  geom_point()
names(p)
p
```


## Data

\small

- The data must be a data frame
- A **copy** of the data is stored in the plot
object (changes to the source dataframe do not change plot)
- Possible to change the data of a plot
object with `%+%`, as in

\scriptsize

```{r}
set.seed(123)
subdiamonds <- sample_n(diamonds,size=100)
p <- p %+% subdiamonds
p
```

## Different data in different layers

\small

- Can specify data for a layer to use.

\scriptsize

```{r}
p + geom_smooth(data=diamonds, method="lm")
```


## Aesthetic mappings

\small

- We have seen that we can specify a default mapping in 
the initialization, or specific mappings
in the layers
    - Warning: specifying a mapping twice can 
    have unexpected consequences.
    
\scriptsize

```{r}
p + geom_point(aes(color=clarity))   
```

## Exercise

Write R code to produce a scatterplot of `price` versus `carat`,
with different colors for values of `clarity`, using the
`subdiamonds` dataset.

## Setting vs mapping 

\small

- An alternative to mapping aesthetics to 
variables is to set the aesthetic to a constant.
    - Set with the layer _parameter_, rather 
    than mapping with `aes()`
    - E.G., set the color of points 
    on a plot to dark blue

\scriptsize

```{r}
ggplot(subdiamonds,aes(x=carat,y=price)) + geom_point(color="darkblue")
```

* **Exercise:** Redo the above plot with the **mapping** color="darkblue" for the `geom_point()` rather
than the **parameter** color="darkblue".
```

## Grouping

\small

- Many `geoms` in ggplot2 group observations
(rows of the dataframe).
    - E.G., a boxplot of a quantitative 
    variable by a categorical variable groups
    observations by the categorical variable. 
- Default `group` is combinations (interaction)
of all categorical variables in the aesthetic specification.
- If this is not what you want, or if there
are no categorical variables, specify `group` yourself.

## `gapminder` data again: Grouping to plot time series

\small

- For plotting time series (multiple measurements
on each observational unit) we want to group
by observational unit.

\scriptsize

```{r}
ggplot(gapminder,aes(x=year,y=lifeExp,group=country)) + geom_line(alpha=.5)
```

## Different groups on different layers

\small

- To add summaries by continent, we need
to specify different groups on different layers.

\scriptsize

```{r, fig.width=4, fig.height=1.8}
ggplot(gapminder,aes(x=year,y=lifeExp,color=continent)) + 
  geom_line(aes(group=country),alpha=0.2) + 
  geom_smooth(aes(group=continent),se=FALSE)
```

* **Exercise** Redo the above, but remove the mapping in `geom_line()` and 
specify `group=country` in the default mapping. 
What grouping variable is used by `geom_line()`? What grouping variable is 
used by `geom_smooth()`?

## Using `interaction()` to specify groups

\small

- Could do boxplots of life expectancy by year 
**and** continent. (Not recommended -- too busy --
just using for illustration.)

\scriptsize

```{r, fig.width=4, fig.height=2.2}
ggplot(gapminder,aes(x=year,y=lifeExp, 
         group=interaction(year,continent),color=continent)) + 
 geom_boxplot()
```

## Overriding `group` on a layer

\small

- Boxplots of life expectancy by year 
and continent on one layer, smoother by 
continent alone on another. 

\scriptsize

```{r, fig.width=4, fig.height=2.2}
ggplot(gapminder,aes(x=year,y=lifeExp, group=interaction(year,continent),
                     color=continent)) + geom_boxplot() +
  geom_smooth(aes(group=continent), se=FALSE)
```

## Geoms

\small

- These are the shapes you want on the plot.
    - See the list of geoms on the cheatsheet.
- Each has a set of aesthetics that are required
for drawing and a set of aesthetics that 
it understands.
    - E.G., `geom_point()` requires x and y position, 
    and understands color, size and shape.
- Aesthetics can also be passed to geoms as parameters.
    - Recall difference between `geom_point(color="darkblue")` and `geom_point(aes(color="darkblue"))`
- Each geom also has a default statistic (stat)
and positional adjustment.
    - More on these next.

## Stats

\small

- Stats are statistical summaries of groups
of data points. E.G.,
    - `stat_smoother()` is a moving average
    of the y positions over windows of x positions.
    - `stat_bin()` is a binning of a quantitative 
    variable into bins of a given width
    - See the cheatsheet for a list.
- Stats create new variables, and these
variables can be mapped to aesthetics.
    - E.G., `stat_bin()` creates
    the variables `count`, `density` and `x`
    - Enclose derived variable name with `..` 
    to use. 

\scriptsize

```{r, fig.width=4, fig.height=2, warning=FALSE}
p <- ggplot(gapminder,aes(x=lifeExp)) + 
  geom_histogram(aes(y= ..density..))
```


## Position adjustment

\small

- See cheatsheet or "Position Adjustments" section of text for more information.
- Default for most geoms is no adjustment ("identity")
- Adjustment to x and/or y position, such 
as "jitter" can reduce overplotting.
- Boxplots in recent plot of gapminder data
were "dodge"d. 
- Histograms of a continuous variable 
by a categorical grouping variable are
"stack"ed by default.

\scriptsize

```{r,fig.width=4, fig.height=1.5, message=FALSE}
gdat <- filter(gapminder,year==1952)
ggplot(gdat,aes(x=lifeExp,color=continent)) + geom_histogram()
```


## Tools for working with layers

- Many possible topics
- In the interest of time, focus on a few 
    - displaying distributions
    - adding error bars and other measures of 
    uncertainty
    - annotating a plot

## Displaying distributions

- The standard histogram is `geom_histogram()`.
    - Displays counts by default, but we have
    seen how to display as density
    - Density is better for comparing the
    shape of distributions
- To include group information, can stack bars
(see previous example),
use faceting to produce separate histograms,
or superpose density histograms.

## Histograms with faceting

\scriptsize
```{r, fig.width=4, fig.height=3}
gdat <- filter(gdat,continent != "Oceania")
h <- ggplot(gdat,aes(x=lifeExp)) 
h + geom_histogram(aes(y= ..density..), binwidth=5) + facet_grid(continent ~ .)
```

## Histograms superposed

\scriptsize


```{r, fig.width=4, fig.height=3}
h + geom_freqpoly(aes(y=..density..,color=continent), binwidth=5)
```

## Density estimation

\small

- `geom_density()` plots a density estimate
    - Think of adding up small normal densities
    centred at each observed datapoint, with 
    the width of each distribution a user-defined
    parameter
- Can superpose multiple density plots.

\scriptsize

```{r, fig.width=4, fig.height=2.2}
h + geom_density(aes(color=continent))
```

## Violin plots

\small

- Instead of superposing, dodge density 
estimates with a violin plot.
    - Violins are density estimate and
    its mirror image displayed vertically

\scriptsize

```{r, fig.width=4, fig.height=2.2}
h + geom_violin(aes(x=continent,y=lifeExp)) 
```

## Adding measures of uncertainty

\small

- `geom_smooth()` includes pointwise error bands by default.
- For factors can add error bars.

\scriptsize

```{r,fig.width=4,fig.height=2,warning=FALSE}
gfit <- lm(lifeExp ~ continent,data=gdat)
newdat <- data.frame(continent=c("Africa","Americas","Asia","Europe")) 
mm <- data.frame(newdat,predict(gfit,newdata=newdat,interval="confidence"))
ggplot(mm,aes(x=continent,y=fit)) +
  geom_point() +  geom_errorbar(aes(ymin=lwr,ymax=upr))
```

## Measures of uncertainty with stat summaries

\small

- Variety of built-in summaries, or can write your own (not
covered).
- Summarize y for different values of x or bins of x values.

\scriptsize

```{r, fig.width=4, fig.height=2.2}
ggplot(gdat,aes(x=continent,y=lifeExp)) + 
 geom_violin() + # superpose over violin plot
 stat_summary(fun.data="mean_cl_normal",color="red") 
```

## Annotating a plot

\small

- Basic tools for annotating are
    - `geom_text()` and `geom_label()` to add text
    - Geoms such as `geom_abline()` to add line segments (see cheetsheet)
    - `labs()` for adding axis labels,
    titles, and captions
    - `annotate()` to add annotations using aesthetics
    passed as vectors to the function, rather than mapped from a dataframe.
- Can add annotations one at a time or many at a time
    - to add many at a time, create a data frame

## Many annotations

\scriptsize

```{r, fig.width=4,fig.height=2.5}
gm07 <- filter(gapminder, year ==2007)
topOilbyGDP <- c("Kuwait","Guinea","Norway","Saudi Arabia")
gdpOil <- filter(gm07,country %in% topOilbyGDP)
ggplot(gm07,aes(x=gdpPercap,y=lifeExp)) + geom_point() +
  geom_text(data=gdpOil,aes(label=country))
```


