---
title: 'Stat 260, Lecture 8, Working with Strings'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages and datasets

\scriptsize

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
```

## Reading

* Strings with `stringr`: Chapter 11 of printed text, 
Chapter 14 of online text.
    * The text emphasizes regular expressions more than 
    we will in class.
* Working with strings (stringr) cheatsheet at
https://github.com/rstudio/cheatsheets/raw/master/strings.pdf]



## Working with ...

* Fixed, or literal strings, like `fish`:
    - count the number of characters in a string
    - detect (yes/no) or find (starting position) substrings 
    - extract and substitute substrings
    - split and combine strings

* String patterns, like `f[aeiou]sh` (more on patterns,
or regular expressions in a minute):
    - detect, find, extract and substitute

* Use tools from the `stringr` package


## The ``stringr` package

\small

- Character string manipulation in base R has evolved over time
as a bit of a patch-work of tools.
    * The names and functionality of these tools has
    been taken from string manipulation tools in Unix and
    scripting languages like Perl.
    * Steep learning curve for many users.
- The `stringr` package aims for a cleaner interface
for tasks that relate to detecting, extracting, replacing
and splitting on substrings.

## Counting the number of characters

\scriptsize

```{r}
mystrings <- c("one fish", "two fish", "red fish", "blue fish")
str_length(mystrings)
```

## Combining Strings with `str_c()`

\scriptsize

```{r}
str_c(mystrings[1],mystrings[2])
str_c(mystrings[1],mystrings[2],sep=", ")
str_c(mystrings[1],NA,sep=", ")
str_c(mystrings[1],str_replace_na(NA), sep=", ")
str_c(mystrings,collapse=", ")
```

## Subsetting Strings with `str_sub()`

- Specify start and stop.
- If stop greater than number of characters, stop at the 
end of the string.
- If `start` greater than number of characters, return `""`

\scriptsize

```{r}
str_sub(mystrings,1,3) 
str_sub(mystrings,-4,-1) # negative means back from end
str_sub(mystrings,1,10000) 
str_sub(mystrings,9,10000) 
```

## Exercise

* For `demog` as defined in the following code chunk,
    1. extract the substring that represents the gender and age category
    (`u` stands for unknown) from each of the three components,
    2. extract the last four characters of each of the three components,
    3. Combine the three components into one string, separated
    by a plus-sign.

\scriptsize

```{r}
demog <- c("new_sp_f014",
           "new_sp_m1524",
           "new_sp_mu")
```

## Fixed Strings vs Regular Expressions

* Fixed strings are interpreted literally, while regular 
expressions are a language for specifying patterns.
    * For example, "fish" is fixed and matches only "fish", 
    while "f[aeiou]sh" matches to "fash", "fesh", ..., "fush".
* Functions from `stringr` that detect/find/extract/substitute
strings can do so with ether fixed strings or regular 
expressions. 
* We will illustrate these functions with fixed strings first, then
discuss regular expressions.
* The text discusses regular expressions first.


## Detecting substrings with `str_detect()`

\scriptsize

```{r}
pattern <- "red"
str_detect(mystrings,pattern)
mystrings[str_detect(mystrings,pattern)]
pattern <- "fish"
str_detect(mystrings,pattern)
```

\normalsize

* (We will later see that we can specify a more general 
pattern than a fixed string.)

## Finding substring starting position

* `str_locate()` returns the start and stop positions of
the *first* occurance of a string.
* `str_locate_all()` returns the start and stop of *all* occurances.

\scriptsize

```{r}
Seuss <- str_c(mystrings,collapse=", ")
str_locate(Seuss,pattern)
str_locate_all(Seuss,pattern)
#str_locate_all(mystrings,pattern)
```


## Replacing (substituting) substrings 

- Use `str_replace` and `str_replace_all`.

\scriptsize

```{r}
str_replace(Seuss,"fish","bird") # replace first occurance
str_replace_all(Seuss,"fish","bird") # replace all
str_replace_all(Seuss,c("one" = "1","two"=2)) # multiple replacements
```


## Splitting Strings

- Some characters in strings, such as `.`, have a special
meaning (more in a minute). 
One option is to wrap  such patterns in `fixed()` for a fixed string

\scriptsize

```{r}
mystrings <- c("20.50", "33.33")
str_split(mystrings,pattern=".")
str_split(mystrings,pattern=fixed("."))
```


## Working with string patterns: regular expressions

\small 

- Regular expressions (abbreviated regexps) are recipes
used to specify search patterns.
- We use character strings to specify regexps in R. 
- Regular expressions is a complex topic. We'll only 
cover the basics.
    

## A simple pattern with `.`

\small

- To illustrate pattern matching, use a simple pattern
`p.n`, meaning `p` followed
by any any character, followed by `n`.

\scriptsize

```{r}
pattern <- "p.n"
mystrings <- c("pineapple","apple","pen")
str_detect(mystrings,pattern)
```

## Matching Special Characters

* Suppose we want to match a pattern involving `.`
* We need to precede, or "escape" the special by a `\`.
* Unfortunately, `\` is a special for character strings,
so we need to escape it too; that is, we need to type
the character string `"\\."` to represent the regexp `\.`.

\scriptsize

```{r}
pattern2 <- "3.40"
mystrings2 <- c("33.40","3340")
str_detect(mystrings2,pattern2)
pattern2 <- "3\\.40"
str_detect(mystrings2,pattern2)
```


## Splitting, Locating and Extracting with Patterns

\scriptsize

```{r}
pattern
str_split(mystrings,pattern)
str_locate(mystrings,pattern)
```

## 

\scriptsize

```{r}
str_extract(mystrings,pattern)
str_match(mystrings,pattern)
```

## Replacing patterns

\small

- `str_replace` and `str_replace_all`
accept regular expressions; e.g., 
```{r}
str_replace(mystrings,pattern,"PPAP")
```

- The replacement string is literal; e.g.,
```{r}
str_replace(mystrings,pattern,"p.n")
```

## Exercise

* Replace the decimals with commas in the following 
strings.

```{r}
exstring <- c("$55.30","$22.43")
str_replace(exstring,"\\.",",")
```



## Adding `*` and `+` quantifiers to `.`

\small

- The combinations `.*` and `.+` match 
multiple characters.
    - E.G., `f.*n` matches `f` followed by 0 or 
    more characters, followed by `n`.
    - `f.+n` matches `f` followed by **1** or more 
    characters, followed by `n`.
```{r}
mystrings <- c("fun","for fun","fn")
pattern1 <- "f.*n"; pattern2 <- "f.+n"
str_extract(mystrings,pattern1)
str_extract(mystrings,pattern2)
```

## "Greedy" matching with `*`

* The `*` quantifier matches the longest possible string.

```{r}
mystrings <- c("fun","fun, fun, fun","fn")
pattern1 <- "f.*n"
str_extract(mystrings,pattern1)
```

## Numerical quantifiers

* Use `{n}` to require exactly `n` matches,
`{n,}` to require n or more, `{,m}` at most m,
and `{n,m}` between n and m

```{r}
str_extract(mystrings,"f.{6}n")
str_extract(mystrings,"f.{1,13}n")
```

## Anchors

* Regular expressions match any part of a string. 
* Use the "anchor" `^` to restrict a match to the start and 
the anchor `$` to restrict a match to the end of a string.
* 

\scriptsize

```{r}
str_extract(mystrings,"^p")
str_extract(mystrings,"e$")
```

## Exercise

* Create a regular expression that matches words that 
are exactly three letters long.

## Other characters to match

* We have illustrated matching on the pattern `.`, which is
any character.
* Instead we can specify a class of characters to match.


\small

```{r}
pattern4 <- "f[aeiou]*n" 
mystrings <- c("fan","fin","fun","fan, fin, fun",
               "friend","faint")
str_extract(mystrings,pattern4) 
```

##

\small

```{r}
str_extract_all(mystrings,pattern4)
```

## Shorthands for Common Character Classes

* `\d` matches any digit (create with "\\d")
* `\s` matches any whitespace (create with "\\s")
* Use a dash to specify a range of characters; e.g.,
    * `[A-Z]` matches capital letters
    * `[a-z]` matches lower-case letters
    * `[1-9]` matches any digit (and so is the same as `\d`)
* Use the caret to negate: `[^abc]` matches anything
except a, b or c.

## Exercise

* Create a regular expression that matches words
that end in `ed` but not `eed`.

## Alternatives

* The `|` in a regular expression is like the logical OR.

\scriptsize

```{r}
str_replace_all(Seuss,"red|blue","color")
str_replace_all("Is it grey or gray?","gr(e|a)y","white")
```
## Converting Case

* Use `str_to_upper()` to change lower- to upper-case and 
`str_to_lower()` to change upper- to lower-case.


```{r}
str_to_upper(Seuss)
```

