---
title: 'Stat 260, Lecture 2, Data Visualization I'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Data Visualization

* Reading: Chapter 1 of the text.
* The key steps of a data analysis are illustrated
in the following figure from the text:

\begin{center}
\includegraphics[height=2.0in]{program.png}
\end{center}

* We will jump into the middle of this program and 
discuss data visualization, because it is often the 
most interesting.

* We will use the `tidyverse` package `ggplot2` to 
visualize data.

## Loading `ggplot2`

* Before you start, make sure you install the 
`tidyverse` collection of R packages. You can 
do this:
    * With the `Tools -> Install Packages` menu item 
    in RStudio. Type `gapminder, tidyverse` into the
    text box and click `Install`.
    * By typing `install.packages("tidyverse")` into 
    the R console.

* Install once, load every time with `library()`:

\scriptsize

```{r}
library(tidyverse)
```



## Example: Car mileage

* Do cars with big engines use more feul that cars with
small engines?

\scriptsize

```{r}
data(mpg)
head(mpg)
```

* Type `View(mpg)` to view the dataset and `?mpg` for 
details on the variables.

## First `ggplot`

* `displ` is the engine displacement in litres and 
`hwy` is the highway mileage in miles per gallon.
* We can plot `hwy` *versus* `displ` as follows:

\scriptsize

```{r,fig.height=4}
ggplot(data=mpg) + geom_point(mapping = aes(x=displ,y=hwy))
```

* Generally a negative linear trend, though there are 
some cars with large engines that get better-than-expected
mileage.

## Second `ggplot`

* We can plot `hwy` *versus* `displ` with
colors to represent different kinds of cars:

\scriptsize

```{r,fig.height=4}
ggplot(data=mpg) + geom_point(mapping = aes(x=displ,y=hwy,color=class))
```

* The `2seaters` are sports cars, with large engines
but light bodies.

* **Exercise:** Redo the above scatterplot but using the
aesthetic `shape=class` to plot different shapes for 
different kinds of cars.




## Components of our `ggplot`

* Our plot requires a dataset, objects 
to plot for each observation in the dataset, 
and a mapping of the variables
in that dataset to features of the plot.

* In `ggplot(data = mpg)` we use the dataset `mpg`,
* we plot points for each observation in the dataset with `geom_point()`,
* and `mapping = aes(x=displ,y=hwy,color=class)`
maps engine displacement to the x-axis, highway mileage to 
the y-axis and vehicle class to the color of each point.

## Geometric objects

* The geometric objects to plot, or "geoms" are specified
by the functions `geom_XXX()`. 

* Example: scatterplot smooths for each class of car.

\scriptsize

```{r,fig.height=4}
ggplot(data=mpg) + geom_smooth(mapping = aes(x=displ,y=hwy,color=class))
```

## Multiple geoms

* Scatterplot smooths are more often plotted over top of points, 
as a summary of the trend. 
* In the following we specify a default aesthetic mapping
that is used by both the points and smooth.

\scriptsize

```{r}
ggplot(data=mpg, mapping = aes(x=displ,y=hwy,color=class)) +
  geom_point() +
  geom_smooth()
```

## Facets

* Instead of using colors to distinguish car class
on one scatterplot, we can split into multiple scatterplots
or "facets".

\scriptsize

```{r}
ggplot(data=mpg, mapping = aes(x=displ,y=hwy,color=class)) + 
  geom_point() + geom_smooth() +
  facet_wrap(~ class,nrow=2)
```


* (The colors are no longer necessary, but are included
to emphasize how our earlier scatterplot has been split.) 

## Faceting

* The facets by car class were added with `facet_wrap()`.
    * The variable to facet on was given in the
    "formula" `~ class`. Formulas will make more sense
    in the next faceting example.
    * `facet_wrap()` would have put made three rows in 
    this example. We force two with `nrow=2`.

* **Exercise:** Repeat the above example, but omit 
the `nrow=2` argument to `facet_wrap()`.

## Faceting on two variables.

* The variable `drv` is  4-, front- or rear-wheel drive.
* The variable `cyl` is the number of cylinders.

\scriptsize

```{r}
ggplot(data=mpg) + 
  geom_point(mapping = aes(x=displ,y=hwy,color=class)) +
  facet_grid(drv ~ cyl)
```

## `ggplot`  layers

- The `g`'s stand for Grammar of Graphics. 
    - Like English grammar is the way in which words
    are put together to form sentences, a grammar of 
    graphics is a way to put together basic graphical 
    elements to make a graph.
- `ggplot`s are built in layers, comprised of 
**data** a **mapping**, a **geom** and optionally **stats**,
such as a scatterplot smoother.
- The layers are arranged and labelled on the graph
by **scale**s and **coord**s (next lecture).
- The data can also be broken into subsets and displayed
in separate graphs by a **facet** specification.


## Another example: The `gapminder` data

* The `gapminder` dataset contains life expectancy, population
size and GDP per capita for countries throughout the 
world from 1952 to 2007 in seven-year intervals.

\scriptsize

```{r}
library(gapminder)
data(gapminder)
head(gapminder)
```


## Subsetting the `gapminder` data

* Plot life expectancy *versus* GDP per capita for 2007. 
    * Need to subset, or "filter" the data to observations from 2007.

\scriptsize

```{r}
gm07 <- filter(gapminder,year==2007)
ggplot(gm07, aes(x=gdpPercap,y=lifeExp,color=continent)) +
  geom_point()
```


## Transform GDP per capita

* GDP per capita differs by several orders of magnitude
across countries. For such explanatory variables, 
the log scale may be more appropriate.

\scriptsize

```{r}
gm07 <- mutate(gm07,log10GdpPercap = log10(gdpPercap))
ggplot(gm07, aes(x=log10GdpPercap,y=lifeExp,color=continent)) +
  geom_point()
```

## Build a plot by layer

* In this example we build the plot by layer and do
not show the plot until all layers are added.
* Our plot will be for the entire `gampinder` dataset.

* Set the mapping:

\scriptsize

```{r}
gapminder <- mutate(gapminder, log10GdpPercap = log10(gdpPercap))
p <- ggplot(gapminder, aes(x=log10GdpPercap,y=lifeExp,color=continent))
```

## Add the geoms

* Overplotting means we can't tell how many points per area
of the plot. 
* Set a transparency, or "alpha" value to make points
semi-transparent. Then many points will add up to solid, and
few points will show as semi-transparent.

\scriptsize

```{r}
p2 <- p + geom_point(alpha=0.1) 
```

- alpha is the transparency aesthetic, between 0 and 1, best
applied directly to the geom.

## Add statistical transformations

- Statistical transformations or **stats** summarize  the data;
e.g., a scatterplot smoother

\scriptsize

```{r}
p2 + stat_smooth()
```

## Exercise:

* The variable `year` is quantitative, but can still be used as a grouping variable.
Do scatterplots of `lifeExp` versus `log10GdpPercap` with points colored by `year`. 
Add a scatterplot smoother with (i) no grouping variable and (ii) `year` as the grouping variable.
Set the SE to FALSE for your smoothers.

## `ggplot` scales

\small

- The **scales** are mappings from the data to the graphics device
    - domain of `continent` is the five continents, range is
    the hexidecimal of the 
    five colors represented on the graph
    - domain of `lifeExp` is 23.599 to 82.603, range is [0,1],
    which `grid` converts to a 
    range of vertical pixels on the graph.
    - legends and axes provide the inverse mapping

\scriptsize

```{r}
p2
```


## `ggplot` coodinate system

\small

- The coordinate system is another layer in how the data
get mapped to the graphics device. 
    - Usually Cartesian, but could be, e.g., polar coordinates,
    or a transformation.
\scriptsize

```{r}
ggplot(gapminder,aes(x=gdpPercap,y=lifeExp,color=continent)) + 
  geom_point(alpha=0.5) + coord_trans(x="log10")
```


## `ggplot` faceting

\small

- How to break up the data into subsets and arrange
multiple plots on the graphics device.

\scriptsize

```{r, fig.width=4, fig.height=2}
p2 +  facet_grid(continent ~ .)
```

## Why so many components?

\small

- A framework for the components of a graph.
- Gives the user the ability to change indvidual 
components one at a time.
