---
title: 'Stat 260, Lecture 6, Tidy Data'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages and datasets

\scriptsize

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
table1 <- read_csv("lec06table1.csv",
                   col_types=cols(
                     country=col_character(),
                     year=col_integer(),
                     cases=col_integer(),
                     population=col_integer()
                   )
)
```

## Create tibbles used in demos

\scriptsize

```{r}
table2 <- table1 %>% 
  gather(cases,population,key="type",value="count") %>%
  arrange(country,year)
table3 <- table1 %>%
  mutate(rate = paste(cases,population,sep="/")) %>%
  select(-cases,-population)
table4a <- table1 %>%
  select(country,year,cases) %>%
  spread(key=year,value=cases)
table4b <- table1 %>%
  select(country,year,population) %>%
  spread(key=year,value=population)
```

## Reading

* Tidy Data: Chapter 9 of printed text, Chapter 12 of online text.
* Advanced: 
Wickham (2014) [https://www.jstatsoft.org/index.php/jss/article/view/v059i10/v59i10.pdf]
* Data import (readr/tidyr) cheatsheet at [https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf]


## Tidy Data

* In a tidy dataset,
    * each variable has its own column,
    * each observation has its own row, and 
    * each value has its own cell.
* The tibble `table1` that we read in earlier is tidy:

\scriptsize

```{r}
table1
```


## Why is table1 tidy?

* It is never straightforward to answer this. 

* These are WHO data on Tuberculosis cases.
    * The  variables are country, year, number of cases and population
    * The observations are contry/year combinations.
    * Each value is in its own cell.
    
* `country` and `year` describe the observational unit, and so there 
must be one row for each, and there must be variables that record
`country` and `year`.
* `cases` and `population` are what we measure on the observational unit and 
so must be variables.

## Non-tidy data

* There are many ways to be non-tidy.
* **Exercise:** Why are `table2` and `table4a` not tidy?

\scriptsize

```{r}
print(table2,n=6)
print(table4a,n=6)
```

## Why tidy?

* Good statistics (exploratory, visualization, modelling)
requires that we identify the observational unit.
* R is efficient at computing with vectors, so variables as column-vectors
are efficient.
* Tidyverse tools require it; e.g.,

\scriptsize

```{r}
table1 %>% mutate(rate=cases/population*100000)
```

##

\scriptsize

```{r}
table1 %>% group_by(year) %>% summarize(sum(cases)) 
ggplot(table1,aes(x=year,y=cases,color=country)) + geom_point()
```

## Exercises

* Compute `rate` from `table2`.
* Compute `rate` from `table4a` and `table4b`.

## Gathering

* The problem with table 4a is that the cases variable is
split across the columns `1999` and `2000` that must be "gathered" into one. 
* Each value of `cases` needs a label (key) to tell us which year it comes from.

\scriptsize

```{r}
table4a %>% gather(`1999`,`2000`,key=year,value=cases)
```

\normalsize

* **Exercise** Repeat for `table4b`

## Gathering: another example

Billboard Top 100 rankings of songs

\scriptsize

```{r}
bb <- read_csv("billboard.csv")
bb
```

## Structure of the Billboard table

- Columns `year` through `date.peaked` describe the song,
then `x1st.week` through `x76th.week` are the 
chart positions for the first through 76th weeks. 
    - If a song is on the chart for less than 76 weeks, its position is `NA` for any missing weeks.
- Weeks are not variables, they are the time data for the 
time series.

## Tidying the Billboard data

- Main step is to gather the rankings in the different
weeks into a `rank` variable. 
- Before gathering, will select/rename some of the variables.
- After gathering, will create some new variables and 
sort the data frame.

## Select and rename

- Won't need `time` or `genre`. 
    - Recall that `select()` from `dplyr` can use `-` to de-select
- Rename `artist.inverted`
    - Recall that `rename()` from `dplyr` takes arguments of the form
    `newname = oldname`

\footnotesize

```{r}
bb <- 
  bb %>% select(-time,-genre) %>% 
  rename(artist = artist.inverted) 
```

## Gather the weeks into a "long" version of the Billboard data

- Leave each song info variable as-is.
- The "values", are the chart positions.
- The weeks are the "keys" for these values.
- We want to create key-value pairs for each observation.
    - There will be missing values, which we can remove.

## 

\scriptsize
```{r}
bb %>% gather(x1st.week:x76th.week,key=week,value=rank,na.rm=TRUE) %>%
  mutate(week= parse_number(week)) %>% # replace, e.g., x1st.week with 1, ...
  arrange(artist,track,week)
```

## Spreading

* Whereas gathering makes a wide dataset long, 
spreading makes a long dataset wide.
* Spread when observations are split across multiple rows.
* E.G., `table2` has observations for each country/year split across 
two rows:

\scriptsize

```{r}
table2
```


## Spreading `table2`

* The column `type` contains the keys for the values in `count` that
belong in columns `cases` and `population`, respectively.

\scriptsize

```{r}
table2 %>% spread(key=type,value=count)
```

* **Exercise** Select `country`, `year` and `cases` from `table1` and 
use spread to obtain a table with rows for each year
and columns for each country. (Note: such data is **not** tidy.)

## Separating

* `separate()` splits a column on a specified separator, or at
a specified character number.

\scriptsize

```{r}
print(table3,n=4)
table3 %>% separate(rate,into=c("cases","population"),sep="/") %>% print(n=4)
```

\normalsize

* Notice that `cases` and `population` are character.


##

\scriptsize

```{r}
table3 %>% separate(rate,into=c("first","remainder"),sep=1)
```

## Convert type of columns after separating

\scriptsize

```{r}
table3 %>% separate(rate,into=c("cases","population"),sep="/",
                    convert=TRUE)
```

## Missing data

* When we used `gather()` on the Billboard data we set
`na.rm=TRUE` to remove weeks where a given track was
not on the charts:

\scriptsize

```{r}
bb %>% select(track,x23rd.week:x25th.week) %>% print(n=4)
```

\normalsize

* The missings in `bb` are "explicit"; when `gather()` removes them
they become "implicit" (e.g., no row for week 25 for track 4).

## Making implicit missing data explicit

* `spread()` will make implicit missing values explicit if
needed for a row.

\scriptsize

```{r}
stocks <- tibble( year=c(2015,2016,2016), qtr = c(1,1,2), 
                  return = c(1.0,2.0,3.0))
stocks
stocks %>% spread(key=year,value=return) 
```

## Make implicit explicit with `complete()`

* `complete()` creates rows for all combinations of input variables
and fills in missing values where necessary.

\scriptsize

```{r}
stocks %>% complete(year,qtr)
```


## Case Study: WHO TB data

* The `who` dataset comes with `tidyr`. We'll use a related (less tidy)
version from the WHO website.

\scriptsize

```{r}
tb <- read_csv("tb.csv")
```

## Structure of TB table

\small

- First column is 2-letter country code, second is year,
third is number of new cases for that country/year.
- Then come TB counts for different gender/age categories.
    - `new_sp` is "new cases by positive pulmonary smear assay"
    - gender is `m` or `f`
    - two special age categories 0-4, 5-14,
    - age categories 0-14, 15-24, 25-34, 35-44, 45-54, 
    55-65, 65+, unknown (`u`)
- Gender/age columns are not variables, they are data
on the observed units.
- Tidy data would have one row for each country, year, gender
and age category, with a column of counts 


## Tidying the TB data

- Recall structure of the data: country, year, count of new cases, counts of new cases by 
gender/age categories.

\scriptsize

```{r}
names(tb)[1:10]
```

\normalsize

- Main step is to "gather" TB prevalence in the different
gender/age categories into a count variable.
    - Complicated by the coding of gender/age categories
- Before gathering, will remove unneeded variables.


## Remove variables

- Won't need overall count
- Special categories 0-4 and 5-14 overlap with 0-14 so remove
- Age unknown not useful for analysing trends, so remove 

\scriptsize
```{r}
tb <- select(tb,-new_sp, -contains("04"), -contains("514"),
             -new_sp_mu, -new_sp_fu)
tb
```

## Gather counts for demographic groups

\footnotesize

- Create demographic variable `demog` and count variable `count` 
by gathering over all variables except `iso2` and `year`.

```{r}
tblong <- tb %>% 
  gather(new_sp_m014:new_sp_f65,key=demog,value=count,na.rm=TRUE)
tblong
```

## Separate gender from age category.

\scriptsize

- First remove `new_sp_`, then separate result on first column 
```{r}
maxlen <- max(nchar(tblong$demog))
tb <- tblong %>% mutate(demog = substr(demog,8,maxlen)) %>%
  separate(demog, into=c("gender","agecat"),sep=1) 
tb
```
