---
title: 'Stat 260, Lecture 5, Reading Data'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages

\scriptsize

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)
```

## Reading

* Workflow: scripts: Chapter 4 of printed text, Chapter 6 of online text
* Introduction to data wrangling: Part II introduction of printed text, Chapter 9 of online text
* Tibbles: Chapter 7 of printed text, Chapter 10 of online text
* Reading data with readr: Chapter 8 of printed text, Chapter 11 of online text
* Data import (readr/tidyr) cheatsheet at [https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf]


## Tibbles

* In base R, the data structure used to hold data sets is the data frame.
* We can make a data frame from vectors as follows:

\scriptsize

```{r}
dd <- data.frame(x=c(NA,10,1),y=c("one","two","three"))
dd
```

\normalsize

* The tidyverse authors find the default behaviour of data frames to 
be odd, and so implemented an improvement called tibbles:

\scriptsize
```{r}
tt <- tibble(x=c(NA,10,1),y=c("one","two","three"))
tt
```


## data frames to tibbles and back

* data frames can be coerced to tibbles and *vice versa*.

\scriptsize
```{r}
as_tibble(dd)
as.data.frame(tt)
```

\normalsize

## tibble printing

* One difference between data frames and tibbles is how they are printed.

* Printing a data frame: all rows and columns, up to your R session's `max.print`.

* Printing a tibble: the first 10 rows, as many columns as fit the screen,
and the column data types.

\tiny

```{r}
flights
```

\normalsize

## Control printing of tibbles

* To see all rows/columns of a tibble, best to `View()` it.

* But you can also print all rows and columns by setting
`options(dplyr.print_min=Inf)` and `options(tibble.width=Inf)`.

## Extracting columns as vectors

* Use the basic tools `$` and `[[` to extract a variable from
a tibble or data frame:

\scriptsize

```{r}
dd$x  
tt$x
dd[["x"]]
tt[["x"]]
```


## Subsetting: columns

* Using `select()` is the preferred method to subset columns
of a data frame or tibble, but we can also use the more
basic tool  `[`; e.g.,

\tiny
```{r}
tt[,"x"]
tt[,c("x","y")]
dd[,"x"] # returns a vector
dd[,c("x","y")]
```


## Subsetting: rows

* Using `filter()` is the preferred method to extract 
rows of a data frame or tibble, but we can also use `[`.

\tiny
```{r}
tt[2,]
tt[1:2,]
dd[2,] 
dd[1:2,]
```

## Exercise

* Create a data frame `myd` and tibble `myt` that each have 
columns named `cat`, `dog` and `mouse`. Each column should be
of length three, but the values in each column are up to you.
* What do `names(myd)` and `names(myt)` return?
* Create the variable `a1 <- c("cat","dog","bird","fish")` and
the variable `a2 <- c("cat","tiger")`. We can combine logicals
with `[` to subset. What do the following return?
    * `myd[,names(myd) %in% a1]`
    * `myd[,names(myd) %in% a2]`
    * `myt[,names(myt) %in% a1]`
    * `myt[,names(myt) %in% a2]`


## Importing data

* We read in the HIV prevalence data with the base R function `read.csv()`,
which returned a data frame.

* We will now discuss the tidyverse equivalent, `read_csv()`,
which returns a tibble.

\scriptsize
```{r, message=TRUE}
hiv <- read_csv("../Labs/HIVprev.csv")
```


## Why use `read_csv()` instead of `read.csv()`?

* `read_csv()` reports how each column of the CSV file was "parsed" (more on this later),
* returns a tibble, 
* uses `stringsAsFactors = FALSE` as the default,
\tiny (recall `hiv <- read.csv("../Labs/HIVprev.csv",stringsAsFactors = FALSE)`) \normalsize
* is faster, and
* is more consistent across operating systems.


## Other `read_` functions

* CSV stands for comma-separated files, aka comma-delimited files
* `read_csv2()` reads semicolon-delimited files,
* `read_tsv()` reads tab-delimited files,
* `read_delim()` reads files with user-specified delimiter.

* Exercise: A file called "chicken.C" contains the following data on two 
chickens, with IDs 22 and 33, who laid 2 and 1 eggs, respectively.
\scriptsize (Reference: https://isotropic.org/papers/chicken.pdf) \normalsize
How would you read this data file into R? 

\scriptsize
```
IDCeggs
22C2
33C1
```

## Skip and comments

* Some files contain a header that describes the data,
aka meta-data, that we should skip when reading.
* Some files include comments that start with
common characters, such as "#".
* Example file

\scriptsize
```
This is a header
that you should skip
# this is a comment
A,B,C
1,2.2,1999-05-10
4,5.5,2001-04-04 # another comment
```

## 

\scriptsize

```{r,message=TRUE}
read_csv("lec05exfile.csv",skip=2,comment="#")
```

## Parsing a vector

* `read_csv()` returns a message that described how
each column of the input file was parsed.
* Parsing a file depends on the `parse_*` functions, such as 
`parse_number()`, that parse vectors. 
* The `parse_*` functions take a vector of character strings as input and
return a vector of a given mode, handling missing
values as specified by the user.

\scriptsize

```{r}
parse_number(c("$10.55","33%","Number is 44","."),na=".")
```

\normalsize

* The parse functions are designed to handle data
formats and character sets from around the world. 
* In this course we assume North American data
formats and character set.
* See the text if you need other formats.

## Other parsing functions

* `parse_logical()`,  `parse_integer()`, `parse_double()`, `parse_character()`,
`parse_factor()`, `parse_datetime()`, `parse_date()` and `parse_time()`.
* Use the `str()` function to see the mode of an object:

\scriptsize

```{r}
str(parse_logical(c("TRUE","FALSE")))
str(parse_logical(c("1","0")))
str(parse_integer(c("1","0")))
str(parse_double(c("1","0")))
str(parse_factor(c("1","0")))
```

## Dates and times

* These parsers have default formats for dates and times,
but your best bet is to specify the format yourself.
* The formatting rules are described in `help(strptime)`.

\scriptsize

```{r}
dd <- c("05/14/1966/12/34/56","04/02/2002/07/43/00","08/17/2005/07/22/00","08/12/2008/16/20/00")
dd <- parse_datetime(dd,format = "%m/%d/%Y/%H/%M/%S")
str(dd)
mean(dd)
diff(dd)
```

## Parsing files

* `read_csv()` and other read functions guess at the format of each
column. Sometimes this works, sometimes not.
* You can read about how these functions guess in the text.
* Here we'll focus on manually specifying the format.

\scriptsize

```{r,message=TRUE}
dat <- read_csv("lec05exfile.csv",skip=2,comment="#")
```

##

* Cut-and-paste the guess and replace parsers as necessary

\scriptsize

```{r,message=TRUE}
dat <- read_csv("lec05exfile.csv",skip=2,comment="#",
                col_types=cols(
                  A = col_integer(),
                  B = col_double(),
                  C = col_date(format = "%Y-%m-%d")
                )
)
str(dat$A)
```

\normalsize

* For reproducibility your R scripts should have a manual
specification of the parsing of each column, rather than
relying on guesses that can change as your data changes.

## Exercise

* Copy the following data to a file and read it in to R.
Specify the column types yourself,
based on the descriptions in the header of the
file. Hint: read about `col_factor()`.

\scriptsize

```
# Variable fert is a factor that records the
# type of fertilizer used in the experiment,
# date records the date and time of the experiment 
# and yield is the yield of corn.
fert,date,yield
F1A2,2018/04/01/12/30,22.56
F1A1,2018/04/02/12/00,26.06
F2A2,2018/04/01/12/45,32.03
F2A1,2018/04/02/12/00,33.21
```

