---
title: 'Stat 260, Lecture 4, Data Transformation'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages

\scriptsize

```{r, message=FALSE, warning=FALSE}
library(gapminder)
library(dplyr)
library(nycflights13)
head(flights)
```

## Reading

* Workflow basics: Chapter 2 of printed text, Chapter 4 of online text
* Data transformation: Chapter 3 of printed text, Chapter 5 of online text
* dplyr cheatsheet at [https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf]


## Data manipulation with `dplyr`

* We have seen the need to manipulate datasets when studying visualization:

\scriptsize
```{r}
gapminder <- mutate(gapminder, log10GdpPercap = log10(gdpPercap))
gm07 <- filter(gapminder,year==2007)
```

\normalsize

* In this lecture we will cover data manipulation more systematically

## `dplyr` 

* The `tidyverse` package `dplyr` contains intuitive tools for 
manipulating data sets.
* These tools are named after verbs.
* The five key verbs (functions) are
    1. `filter()` to select subsets of observations (rows).
    2. `arrange()` to reorder rows
    3. `select()` to select variables (columns)
    4. `mutate()` to create new variables from existing ones, and
    5. `summarize()` to calculate summary statistics

## Filter rows with `filter()`

* We previously used `filter()` to extract a subset of the 
`gapminder` dataset based on a logical condition.

\scriptsize
```{r}
gm07 <- filter(gapminder,year==2007)
```


* `gm07` now contains the rows of `gapminder` in which the variable
`year` is equal to 2007.
* The `==` is a comparison. 
* We specify the rows to extract with comparison and logical operators.

## Relational (comparison) operators

\scriptsize

* The basic relational operators are described in `?Comparison`:
`==` is equal, `!=` is not equal, `>` is greater than, `<` is
less than, `>=` is greater than or equal, `<=` is less than or equal.
* Watch out for finite-precision arithmetic.

```{r}
2>3; 2/2 == 1; sqrt(2)^2 == 2; near(sqrt(2)^2,2) 
```

## Logical operators

\scriptsize

- The basic logical operators are described in `?Logic`: 
`!` is NOT, `&` is AND, `|` is OR.
```{r}
!TRUE ; TRUE | FALSE & FALSE; (TRUE | FALSE) & FALSE # eval parentheses first
```

- Relational comparisons can be combined with logicals
```{r}
(2==2) | (2==3) 
```

- Logical operations between vectors are element-wise.
```{r}
x <- c(TRUE,TRUE,FALSE); y <- c(FALSE,TRUE,TRUE)
!x ; x&y ; x|y
```

## `filter()` example

* Extract all flights from January, with departure delay of 1 or more:

\scriptsize

```{r}
jan13 <- filter(flights, month==1 & dep_delay > 1)
```

\normalsize

* **Exercises** Extract all flights from January or February.
Extract all flights from January or February that have 
a departure delay of 1 or more.




## Many logicals and `%in%`

* In lab 2 we saw the `%in%` operator:

\scriptsize

```{r}
hiv <- read.csv("../Labs/HIVprev.csv",stringsAsFactors = FALSE)
cc <- c("Botswana","Central African Republic","Congo","Kenya","Lesotho","Malawi","Namibia","South Africa","Swaziland","Uganda","Zambia","Zimbabwe")
hihiv <- filter(hiv,Country %in% cc)
```

\normalsize

* This is a convenient alternative to 

\scriptsize

```
filter(hiv,Country=="Botswana" | Country=="Central African Republic", etc. )
```

\normalsize

* **Exercise** The `nycflights13` package includes a table `airlines` that
translates the two-letter airline codes in `flights` into the 
full names of the airlines. Extract all flights operated by United, American
or Delta.



## Missing data: `NA`

* In R, missing data (not available) is denoted by `NA`.
* `NA` takes precendence in all comparison and arithmetic operations,
and almost all logical operations.

\scriptsize

```{r}
NA > 3; NA+10; NA & TRUE; NA | TRUE
```


## `is.na()`

* Test for `NA` with `is.na()`:

\scriptsize

```{r}
vv <- c(10,NA,1)
is.na(vv)
vv>1
is.na(vv) | vv> 1
```

\normalsize

* **Exercise** Extract all flights from January with missing departure delay.


## Sorting with `arrange()`

* `arrange()` changes the order of rows, putting `NA`s last

\scriptsize

```{r}
vv <- tibble(x=c(NA,10,10,1),y=c("one","two","three","four"))
arrange(vv,x)
arrange(vv,x,y)
```

## Descending order

* The default in `arrange()` is increasing order.
* Use `desc()` for descending order.

\scriptsize

```{r}
arrange(vv,desc(x))
```

\normalsize


* **Exercise** Arrange the `mpg` data set by *decreasing order* in the 
number of cylinders (variable `cyl`) and *increasing order* 
by engine displacement (variable `displ`)
within cylinders. 


## Selecting columns with `select()`

* Where as `filter()` subsets by row, `select()` subsets by column. 
    * We specify columns by their name, possibly using helper functions.

* Select month, day, hour and minute from `flights`:

\scriptsize

```{r}
select(flights,month,day,hour,minute)
```

## Ranges

* Select or de-select a range of columns with `:`


\tiny

```{r}
select(flights,month:minute)
select(flights,-(month:minute))
```

## Renaming variables

* You can rename variables as they are selected.
* To rename a variable without selecting it use `rename()`.

\scriptsize

```{r}
flights <- rename(flights,tail_num = tailnum)
```

## Helper functions

* Some useful helper functions are `starts_with()`, `ends_with()` and `contains()`.
    * See `?select` for a complete list.
    
\scriptsize

```{r}
select(flights,contains("dep"))
```

* **Exercise** Select all variables with "dep" or "arr" in the name.

## Add new variables with `mutate()`

* We've used `mutate()` to add log-GDP to the `gapminder` data set.

\scriptsize

```{r}
gapminder <- mutate(gapminder, log10GdpPercap = log10(gdpPercap))
```

\normalsize

* You can use variables as they are created by `mutate()`:

\scriptsize

```{r}
flights <- mutate(flights,gain = arr_delay - dep_delay, gainh = gain/60)
```

## Summaries and grouping

* A common task is "split-apply-combine": We want to 
split a data set into groups defined by one variable, apply a statistical
summary to each group, and then combine the results.
* With `dplyr` we use `group_by()` and `summarize()`.

\scriptsize

```{r}
by_day <- group_by(flights,month,day)
summarize(by_day,count = n(), delay = mean(dep_delay,na.rm=TRUE))
```

\normalsize

* Note: `mean()` returns `NA` whenever
there are missing values, unless we set `na.rm=TRUE`, 

## Combining operations with the pipe

* The pipe `%>%` can be used to chain together operations,
without the need to store intermediate results

\scriptsize

```{r}
select(flights,month,day,dep_delay) %>% 
  group_by(month,day) %>% 
  summarize(count = n(), delay=mean(dep_delay,na.rm=TRUE)) 
```

\normalsize

* Notice that we omit the data set name when a function receives
data from the pipe.
