---
title: 'Stat 260, Lecture 7, Relational Data'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages and datasets

\scriptsize

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)
# Example from http://www.itl.nist.gov/div897/ctg/dm/sql_examples.htm
station <- read_csv("station.csv",col_types=cols(ID=col_integer()))
stats <- read_csv("stats.csv",col_types=cols(ID=col_integer()))
```

## Reading

* Relational Data: Chapter 10 of printed text, Chapter 13 of online text.
    * In the sections on "Joins", we will focus on 
    the left-join (in the **Mutating Joins** section)
    and the semi-join (in the **Filtering Joins** section).
* Data transformation (dplyr) cheatsheet at
[https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf]


## Multiple Tables

* Modern data comes in multiple tables, called relational data. 
* Such structure is motivated by relational database management
systems (RDBMS) that revolutionized database management.
* **Example**  `station` and `stats` tables for weather data:

\scriptsize

```{r}
station
stats
```

## Relations in the Weather Data

* The relation, or connection between `station` and `stats` is the
`ID` variable present in both. 
* Think of the `ID`s as short-hand for the info in the `station` table.
    * For example, `ID` 13 from `station` is short-hand for:
    Phoenix AZ at latitude 33N and longitude 112W.  
* The `stats` table is much more concise for not repeating the 
info on each station.

## Joining tables

* However, in some cases we may wish to include information, such as station 
name, in the `stats` table. 
    * The text calls this a "mutating join".
* Or, we may wish to filter weather measurements in `stats` based on
the characteristics of the stations, such as latititude `Lat_N >= 40`.
    * The text calls this a "filtering join".

## The `nycflights13` Relational Data

* We used the `flights` data from `nycflights13`. There are several other 
tables in this package:

\scriptsize

```{r}
print(airlines,n=3)
print(airports,n=3)
```

##

\scriptsize

```{r}
print(planes,n=3)
print(weather,n=3)
```

## Relations in the `nycflights` Tables

* Figure from Chapter 10 of R for Data Science:

\vspace*{-.2in}

\begin{center}
\includegraphics[height=2.5in]{lec07relatdat.png}
\end{center}

* **Exercise:** The relationship between the `weather` and `airports` tables
is not shown on the diagram. What is it?

## Keys

* The `ID` variable in the `station` table is its "primary key". 
* It uniquely identifies observations (rows)
in the table.
* The `ID` column in the `stats` table is a "foreign key"
that links to the `station` table.
* Primary/secondary keys are the "relations" in relational data.

\scriptsize

```{r}
print(station,n=3)
print(stats,n=3)
```

## Multiple Keys

* It may take multiple variables to uniquely identify an observation
in a table.
* For example, in the `weather` table from `nycflights13`, 
we need `year-month-day-hour` plus `origin`.

\scriptsize

```{r}
print(weather,n=3)
```

## Tables with no Primary Key

* Some tables lack the variables needed to uniquely identify 
its observations.
* For example, in the flights table `year-month-day`, `flight` and
`tailnum` do **not** identify the flight.

\scriptsize

```{r}
flights %>% 
  count(year,month,day,flight,tailnum) %>%
  filter(n > 1) %>% print(n=3)
```

## Surrogate Key

* You can add a "surrogate key" to a table with no 
primary key using `mutate()` and  `row_number()`.

* **Exercise** Select year, month, day, flight and tailnum
from `flights` and add a surrogate key to this 5-column table.

## Joining Tables

* To start, add the data in our
`stations` table to the weather statistics
in the `stats` table.
    * This is what the text calls a mutating joing,
    because it adds columns to `stats` like a 
    call to `mutate()` would.

\scriptsize

```{r}
stats %>% left_join(station)
```

## A Note on Joins

* The term "join" is from SQL (Structured Query Language),
which is the standard language used to construct
RDBMS queries.
* There are **many** types of joins. In this class
we will focus on the two that I think are most
useful in data analysis, the left-join (a mutating join)
and the semi-join (a filtering join).
* However, for a bit of context, we will
discuss inner- and outer-joins.

## Inner Joins

* To illustrate inner- and outer-joins we
remove one of the rows of `station`.

\scriptsize

```{r}
station <- station[-3,]
```

\normalsize

* Now repeat the `left_join()`:

\scriptsize

```{r}
stats %>% left_join(station)
```


## The Left-Join as an Outer Join

* In the output of `left_join()`, the data from `stats` is
on the left.
* The `left_join()` is an "outer join", because
it keeps observations in one or more of the tables,
in this case the left table.
* Similarly, a right-join keeps all observations
in the right table, and a full-join keeps all 
observations in both tables.

## The Inner-Join

* The inner-join keeps observations that appear
in **both** tables.

\scriptsize

```{r}
stats %>% inner_join(station)
```

\normalsize

* This might be good if we only wanted data from
stations in `station`, but there is a tendency 
to accidentally loose data.
    * Better to use a filtering join (next topic).

## Defining the Key Columns

* The `station` and `stats` tables are very easy 
to join because the primary key in `station` has
the same name as the primary key in `stats`.
* The `by` argument to `left_join()` lets you 
specify the keys to match on.
* The default is `by = NULL`. 
    * Uses all variables that appear in both tables; 
    `ID` for `station` and `stats`.
    * This is called a "natural" join.

## Left-Joins with `nycflights13`

* To illustrate left-joins with the `nycflights13` data
we use the reduced `flights` table defined in the text.

\scriptsize

```{r}
flights2 <- flights %>%
  select(year:day, hour, origin, dest, tailnum, carrier)
flights2
```

## Natural Join of `weather`

* `flights2` and `weather` share the variables
`year`, `month`, `day`, `hour` and `origin`.

\scriptsize

```{r}
flights2 %>% left_join(weather)
```

## `by = x`

* Use `by=x` to join on a specific column. 
* For example, `year` means something different
in `flights2` and `planes`. Use only `tailnum`.

\scriptsize

```{r}
flights2 %>% left_join(planes, by="tailnum")
```

## Matching Keys with Different Names

* The airport codes are in either `origin` or `dest` in 
the `flights2` table, and in `faa` in the `airports`
table. 
* Use, e.g., `by= c("dest" = "faa")`

```{r}
flights2 %>% left_join(airports,by=c("dest" = "faa"))
```

## Exercise

* Change the name of the `ID` column in `stats` to `Station`.
* With these modified tables, do a left-join of the
`stats` and `station` tables.

## Filtering Joins

* In past lectures we have use `%in%` for filtering
a table according to a character string.
    * E.G., `gapminder %>% filter(country %in% c("Canada","United States"))`
* Filtering joins are an extension 
to filter a table according to another table.

\scriptsize

```{r}
top_dest <- flights2 %>% count(dest,sort=TRUE) %>% head(n=10)
print(top_dest,n=4)
```


##

* The "old" way and the semi-join way:

\scriptsize

```{r}
flights2%>% filter(dest %in% top_dest$dest) %>% print(n=4)
flights2 %>% semi_join(top_dest) %>% print(n=4)
```


## Notes

* The `n` column of the `top_dest` table used
for filtering does not appear in the output. 
* This ensures the same behaviour as the "old" way.
* The power of `semi_join()` is in matching 
to multiple columns, as in the following exercise.

## Exercise

* From the original `flights` table,
create a table called `top_dep_delay`
comprised of the year-month-days with the 
3 largest total delays, where total delay
is defined as the sum of the `dep_delay` variable
for each year-month-day. 
    * Hints: use `group_by()` to group
    `flights` by year-month-day; use `summarize()` to
    compute total delays (watch out for missing values);
    use `arrange()` to sort on your total delays variable
    (you want to sort in descending order)
* Do a semi-join 
to filter `flights` to these days.


## Set Operations

* An add-on to this chapter is the set operations, 
`intersect()`, `union()`, and `setdiff()`, 
which can act on pairs of tables.
* Illustrate with `intersect()`.

\scriptsize

```{r}
v1 <- c("apple","pen","pineapple"); v2 <- c("apple","orange","grape")
intersect(v1,v2)
df1 <- tibble(x=c(1,2),y=c(1,1)); df2 <- tibble(x=c(1,1),y=c(1,2))
intersect(df1,df2)
```

## Case Study: TB Map

\scriptsize

```{r}
tb <- read_csv("tb.csv") %>%
  select(-new_sp, -contains("04"), -contains("514"),
             -new_sp_mu, -new_sp_fu)  %>%
  gather(new_sp_m014:new_sp_f65,
         key=demog,value=count,na.rm=TRUE)%>%
  mutate(demog = substr(demog,8,12)) %>%
  separate(demog, into=c("gender","agecat"),sep=1) 
tb2 <- tb %>% spread(key=gender,value=count) %>%
  mutate(count = m+f) %>% 
  select(-m,-f) %>%
  filter(!is.na(iso2))
```

##

* Use map data from the `maps` package

\scriptsize

```{r}
library(maps) # install.packages("maps") first
ww <- as_tibble(map_data("world")) %>%
  mutate(iso2 = iso.alpha(region))
ww
```

##

* Extract TB data on children in 2000 and join
to `ww`.

\scriptsize

```{r}
tbchild00 <- tb2 %>% 
  filter(agecat == "014",year==2000) %>% 
  select(iso2,count)
wwchild00 <- ww %>% left_join(tbchild00)
```

##

* Can't distinguish countries on the count scale,
so transform to log-count (plus 1 to avoid log of 0).

\scriptsize

```{r}
ggplot(wwchild00,aes(x=long,y=lat,fill=log(count+1),
                     group=group)) + 
  geom_polygon() + coord_quickmap()
```

## Improvements

* Crude numbers of cases are not very informative.
Large countries have large counts.
* Would have been better as rates: 
   * Need a table of the population of children
   in each country in 2000 and the 
   ISO 2 country code.
   * Join population to the TB data, and calculate
   rates per 100,000
   
