---
title: 'Stat 260, Lecture 12, Iteration'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages and datasets

\scriptsize

```{r, message=FALSE, warning=FALSE}
library(tidyverse) 
```

## Reading

* Iteration chapter of R for Data Science, 
chapter 17 of printed book and chapter 21 online.
* purrr cheatsheet [https://github.com/rstudio/cheatsheets/raw/master/purrr.pdf]

## Iterating over a vector

* For loops allow iteration.
* A common scenario for iteration is that our data is in a vector (list), 
and we want to perform the same operation on 
each element.
* Such iteration is so common that special tools have been developed
with the aim of reducing the amount of code (and therefore
errors) required for common iterative tasks.
    - Tools in base R include the `apply()`
    family of functions. 
    - A tidyverse package called `purrr` includes more.
    
## Example data

* To illustrate iteration we simulate data and fit four 
regression models.

\scriptsize

```{r}
set.seed(42)
n <- 100
x1 <- rnorm(n); x2<-rnorm(n)
y1 <- x1 + rnorm(n,sd=.5); y2 <- x1+x2+rnorm(n,sd=.5)
y3 <- x2 + rnorm(n,sd=.5); y4 <- rnorm(n,sd=.5)
rr <- list(fit1 = lm(y1 ~ x1+x2),
  fit2 = lm(y2 ~ x1+x2),
  fit3 = lm(y3 ~ x1+x2),
  fit4 = lm(y4 ~ x1+x2))
coef(rr$fit1)
```

## Exercise

* The elements of the list `rr` from last slide are
`lm` objects. The function `coef()` is generic.
Assign class "lm_vec" to `rr` and write a 
`coef()` method for objects of this class.
Hint: Your function could encapsulate 
a `for()` loop like the following.

```{r}
for(i in seq_along(rr)) { # safer than 1:length(rr) 
  coef(rr[[i]])
}
```


## Extracting the regression coefficient for `x1`

* Using a `for()` loop, we initialize an object to hold the 
**output**, 
loop along a **sequence** of values for an index variable, 
and execute the **body** for each value of the index variable.

\small

```{r}
betahat <- vector("double",length(rr))
for(i in seq_along(rr)) { # safer than 1:length(rr) 
  betahat[i] <- coef(rr[[i]])["x1"]
}
betahat
```

## Looping over elements of a set

* The index set in the `for()` loop can be general. 
    - We might use this generality to loop over named components of a list.
    
\small

```{r}
fits <- paste0("fit",1:4)
for(ff in fits) {
  print(coef(rr[[ff]])["x1"])
}
```

\normalsize

* Looping over a set makes it harder to save the results, though.

## Avoid growing vectors incrementally

\scriptsize

```{r}
means <- seq.int(1000)
set.seed(123)
system.time({
output <- double()
for (i in seq_along(means)) {
  n <- sample(100, 1)
  output <- c(output, rnorm(n, means[[i]]))
}
})
```

##

\scriptsize

```{r}
system.time({ 
out <- vector("list", length(means))
for (i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}
out <- unlist(out)
})
```


## The body of a loop can be a small part of the code

* In our examples, most of the code is for setting up the 
output and looping, with very little to do with the body.

* To illustrate, consider a small change: instead of the estimated coefficient of `x1` we wanted
the estimated coefficient of `x2`:

\small

```{r}
betahat <- vector("double",length(rr))
for(i in seq_along(rr)) { # safter than 1:length(rr) 
  betahat[i] <- coef(rr[[i]])["x2"]
}
betahat
```

## Exercise

* Write a `for()` loop to find the `mode()` of
each column in nycflights13::flights


## Using `lapply()`

* The intent of `lapply()` is to take care of 
the output and the loop, allowing us to focus 
on the body.

\scriptsize

```{r}
b1fun <- function(fit) { coef(fit)["x1"] } # body
bfun <- function(fit,cc) { coef(fit)[cc] } # body
lapply(rr,b1fun)  # or sapply(rr,b1fun) or unlist(lapply(rr,b1fun))
lapply(rr,bfun,"x1")  
```


## Exercise

* Re-write your summary method for objects of 
class `lm_vec` to use `lapply()`.

## Iterating with the `map()` functions from `purrr`

* The `purrr` package provides a family of functions 
`map()`, `map_dbl()`, etc. that do the same thing
as `lapply()` but work better with other tidyverse functions.
    * `map()` returns a list, like `lapply()`.
    * `map_dbl()` returns a double vector, etc.

\scriptsize

```{r}
library(purrr)
map_dbl(rr,b1fun) # or rr %>% map_dbl(b1fun)
# map_dbl(rr,bfun,"x1")
```

## Exercises

* Use `map_chr()` to return the `mode()` 
of each column of the `nycflights13::flights`
tibble.
* Use `map()` to return the `summary()`
of each column of the `nycflights13::flights`
tibble.

## Pipes and `map()` functions

* Suppose we want to record a model summary returned by 
the `summary()` function.
    * `summary()` applied to
    an `lm()` object it computes regression summaries like
    standard errors and model R$^2$.

\small

```{r}
rr %>%
  map(summary) %>%
  map_dbl(function(ss) { ss$r.squared })
```

\normalsize

## 

* Notice that we can define a function on-the-fly in the
call to a `map()` function.

* `map()` functions have a short-cut for function
definitions.

\small

```{r}
rr %>%
  map(summary) %>%
  map_dbl(~.$r.squared)
```

\normalsize

* In `~.` read `~` as "define a function" and `.` as "argument to 
the function"

## Exercise

* Re-write the code `map_dbl(rr,b1fun)`
with the forward pipe, as in the previous 
slide, and define the function `b1fun` on the fly.
Use the `~` and `.` to define your function.

## Detour: The apply family of functions in R

\small

- The "original" apply is `apply()`, which can 
be used to apply a function to rows or columns 
of a matrix.

\scriptsize

```{r}
mat <- matrix(1:6,ncol=2,nrow=3)
mat
apply(mat,1,sum) # row-wise sums; rowSums() is faster
apply(mat,2,sum) # column-wise; colSums() is faster
```

## Detour, cont.

\small

- `sapply()` takes the output of `lapply()` and simplifies
to a vector or matrix.

\scriptsize

```{r,eval=FALSE}
FIX -- this does not work
fsum <- function(x) { sum(x$FTEs) }
sapply(sp.stat,fsum)[1:2]
````

## Detour, cont.

\small

- Other `apply`-like functions `vapply()`, `mapply()`,
`tapply()`, ...
- I don't use these.
    - See their respective help pages for information.
    
