---
title: 'Stat 260, Lecture 10, Programming, Pipes and Functions'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages and datasets

\scriptsize

```{r, message=FALSE, warning=FALSE}
library(tidyverse) # loads the %>% pipe
```

## Reading

* Programming overview, Pipes and Functions: Intro to Part II, and Chapters 14 and 15 of the printed text,
Chapters 17 to 19 of online text.
* Optional: Wickham (2014), Advanced R, Chapter 6,
[http://adv-r.had.co.nz/Functions.html]

## Programming in R

* We have already been programming in R, mostly by writing
chunks of code that use tidyverse functions to do
data visualization or wrangling.
* Now we discuss strategies for making our 
code easier to read and less prone to errors.
* One useful tool for this purpose is the forward pipe,
`%>%`, which we'll discuss in more detail.
* A principle in programming is "Do not Repeat Yourself (DRY)",
and writing functions can help us stay DRY.

## Pipes

* The forward pipe `%>%` is implemented in the 
`magrittr` package, and is loaded when we load the tidyverse.
* Forward pipes are useful for combining a linear
sequence of data processing steps, 
when we won't need the intermediate steps.
    * Tidyverse functions are typically named as actions, or
    verbs, and a linear sequence of such actions 
    reads like a sentence.
    
## We can use pipes too much

* However, we can take the idea too far and
make the code difficult to debug.

* **Exercise** Debug the following code

\scriptsize

```{r,eval=FALSE}
read_csv("API_ILO_country_YU.csv") %>%
  gather(year,`Unemployment Rate`,`2010`:`2014`,convert=TRUE) %>%
  filter(str_detect(`Country Name`," \\(IDA.*\\) %>%
  mutate(`Country Name` = 
           str_replace(`Country Name`,"", "\\(IDA.*\\)")) %>%
  select(-`Country Code`) %>%
  ggplot(aes(x=year,y=`Unemployment Rate`, color=`Country Name`)) %>% 
  + geom_point() + geom_line()
```

## Pipes are not good when there are multiple inputs

* Code may involve
parallel computations that are assembled at the end.
* For example, suppose you need to read in two tibbles, 
manipulate each with actions like filter/gather/mutate,
and then join them together. 
    * We should save the tibbles to intermediate objects
    before joining them.

## Other Tools from  `magrittr` 

* See the text's section -- I don't use them myself
* Note: `%$%` is similar to the base R function `with()`

\scriptsize

```{r}
library(magrittr)
mtcars %$% cor(disp,mpg)
with(mtcars,cor(disp,mpg))
```


## R functions -- overview

* Encapsuling code in a function has several advantages:
    * can be used multiple times on different inputs
    * can compartmentalize computations and give them a name
    * can help you break a complicated task down into 
    more manageable pieces.
* We will discuss: 
    * when to write a function, 
    * components of a function
    * writing pipeable functions,

## When to write a function

* If you find yourself cutting and pasting the same
code multiple times (more than twice, according
to the text), then you should consider writing a function.
* See text for one example. Here is another.
* The `Boston` dataset in the `MASS` package includes
data on house prices (`medv`) and characteristics 
of different neighborhoods in Boston
[https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html]
* Certain kinds of statistical analyses of the
relationship between `medv` and the other
variables require that the other variables be standardized,
by subtracting the mean values and dividing by the SD.

* I have made the dataset available in our Notes folder
with the following

\scriptsize

```{r,eval=FALSE}
library(MASS)
write_csv(Boston,file="Boston.csv")
```


## `Boston` dataset

\scriptsize

```{r}
Boston <- read_csv("Boston.csv")
dim(Boston)
Boston
```

## Standardize columns of `Boston`

* You can standardize the first column of `Boston` with

\scriptsize

```{r}
Boston$crim <- (Boston$crim - mean(Boston$crim, na.rm=TRUE))/
  sd(Boston$crim,na.rm=TRUE)
```

\normalsize

* Now cut-and-paste 12 times to standardize the remaining
predictors of `medv`

\scriptsize

```{r}
Boston$zn <- (Boston$zn - mean(Boston$zn, na.rm=TRUE))/
  sd(Boston$zn,na.rm=TRUE)
Boston$indus <- (Boston$indus - mean(Boston$indus, na.rm=TRUE))/
  sd(Boston$indus,na.rm=TRUE)
# Etc.
```

\normalsize

* Not only is this tedious, it is error-prone. Plus, 
we will need to do the same operation on other datasets.

## A standardization function

* The following function standardizes a vector.
    * We'll learn more about the components of a 
    function in the slides to follow.
    
\scriptsize

```{r}
standardize <- function(x) {
  (x - mean(x,na.rm=TRUE))/sd(x,na.rm=TRUE)
}
Boston$crim <- standardize(Boston$crim)
Boston$zn <- standardize(Boston$zn)
# Etc
```

\normalsize

* This has reduced the amount of code and 
chances for cut-and-paste errors.


## Components of a function

- In R, functions are objects with three essential components: 
    - the code inside the function, or `body`,
    - the list of arguments to the function, and
    - a data structure called an `environment` which is like a 
    map to the memory locations of all objects defined
    in the function.

## Example function 

\scriptsize

```{r}
f <- function(x) {
  x^2
}
f
```


## The function arguments

- These are the arguments to the function. 
- Function arguments can have default values, as in:

\scriptsize

```{r}
f <- function(x=0) { x^2} 
```

\normalsize

* **Exercise** Re-write our `standardize()` function
to have an additional argument `na.rm`, set 
to `TRUE` by default.

## Argument defaults

- Argument defaults can be defined in terms of other arguments:

\scriptsize

```{r}
f <- function(x=0,y=3*x) { x^2 + y^2 }
f()
f(x=1)
f(y=1)
```


## Argument matching 

- When you call a function, the arguments are matched first by 
name, then by "prefix" matching and finally by position:

\scriptsize

```{r}
f <- function(firstarg,secondarg) {
  firstarg^2 + secondarg
}
f(firstarg=1,secondarg=2)
f(s=2,f=1)
f(2,f=1)
f(1,2)
```


## The function body

- This is the code we want to execute. 
- When the end of a function is reached,
the value of the last line is returned.
    - If you prefer, you can end a function with 
    `return()` to signal the function's returned value.
    
\scriptsize

```{r}
f <- function(x=0) { x^2} 
f <- function(x=0) { return(x^2)}
```
    
## Control Flow

* Code within a function is not always executed
linearly from start to end.
* We may need to execute different code chunks 
depending on the function inputs.
* We may need to repeat certain calculations, 
or loop.
* Such constructs are called control flow.
* We'll touch on some of the basics.

## `if` and `if`-`else` 

- `if` tests a condition and executes code if the condition is true. Optionaly, can couple with an `else` to specify code to execute
when condition is false.

\scriptsize

```{r}
if("cat" == "dog") {
  print("cat is dog")
} else {
  print("cat is not dog")
}
```

## Conditions require `TRUE` or `FALSE`

* If not, you will get a warning or error. 

\scriptsize

```{r,eval=FALSE}
if(c(TRUE,FALSE)) {} # Throws the following warning:
#Warning message:
#In if (c(TRUE, FALSE)) { :
#  the condition has length > 1 and only the first element will be used
if(NA) {} # Throws the following error:
#Error in if (NA) { : missing value where TRUE/FALSE needed
```

## `for` loops

* Example: In the following, `1:nreps` is the "index set".

\scriptsize

```{r}
n <- 10; nreps <- 100; x <- vector(mode="numeric",length=nreps)
for(i in 1:nreps) { # or i in seq(nreps)
  x[i] <- mean(rnorm(n))
}
summary(x)
print(i)
```

\normalsize

* **Exercise** Write a function `standardize_tibble()`
that loops through the columns of a `tibble` and 
standardizes each with your `standardize()` function.
Hints: If `tt` is a tibble, `ncol(tt)` is 
the number of columns, and `1:ncol(tt)` is 
an appropriate index set. If `tt` is a tibble,
`tt[[1]]` is the first column.


## `for` loop index set

- Index sets are sometimes the indeces of a vector, 
and can also be the elements of the vector.

\scriptsize

```{r}
ind <- c("cat","dog","mouse")
for(i in seq_along(ind)) {
  print(paste("There is a",ind[i],"in my house"))
}
for(i in ind) {
  print(paste("There is a",i,"in my house"))
}
```

\normalsize

* We'll learn more about this kind of iterating in the chapter
**Iteration with `purrr`**.

## `while` loops

\small

- Use a `while` loop when you want to continue until
some logical condition is met.

\scriptsize

```{r}
set.seed(1)
# Number of coin tosses until first success (geometric distn)
p <- 0.1; counter <- 0; success <- FALSE
while(!success)  {
  success <- as.logical(rbinom(n=1,size=1,prob=p))
  counter <- counter + 1
}
counter
```

## `break`

- `break` can be used to break out of a `for` or `while` loop.

\scriptsize

```{r}
for(i in 1:100) {
  if(i>3) break
  print(i)
}
```


## The function environment

- The environment within a function is like a map to the 
memory locations of all its variables.
- The function arguments are "passed by value", meaning that
a copy is made and stored in the function's environment.
- Variables created within the function are also store in its
environment

\scriptsize

```{r}
f <- function(x) {
  y <- x^2
  ee <- environment() # Returns ID of environment w/in f
  print(ls(ee)) # list objects in ee
  ee
}
f(1) # function call
```

## Enclosing environments

- Our function `f` was defined in the global environment,
`.GlobalEnv`,
which "encloses" the environment within `f`.
- If `f` needs a variable and can't find it whithin
`f`'s environment, it will look for it in the enclosing 
environment, and then the enclosing environment of `.GlobalEnv`,
and so on.
- The `search()` function lists the heirarchy of 
environments that enclose `.GlobalEnv`. 

\scriptsize

```{r}
search()
```

\normalsize

- To facilitate this search, each environment includes 
a pointer to its enclosing environment.

## Exercise 

*  The following code chunk is typed into the R Console.
    - What is the output of the function call `f(5)`?
    - What is the enclosing environment of `f()`?
    - What is the enclosing environment of `g()`?
    - What search order does R use to find the 
    value of `x` when it is needed in `g()`?
```{r}
x <- 1
f <- function(y) {
  g <- function(z) {
    (x+z)^2
  }
  g(y)
}
```

## Writing readable code

* The more your code does, the harder it is for
others to read.
    * Here "others" includes you some time in the future.
    * Wickham says we should write code that future-you 
    can understand, because past-you doesn't answer emails.
* See the **Functions are for Humans and Computers**
section of text for tips on writing readable code.

## Other reasons to write functions

* Functions can be used to prevent repetition, 
but even if used only once they can improve 
code readability.
    * For example, you are writing a function `func()`
    that computes a statistic `mystat` 
    that takes 10 lines of code to calculate. 
    * The rest
    of your function is only 5 lines. 
    * Write a function
    called `mystat()` and call it from `func()`.
    * If you define `func()` first, it will be 
    easier to document `mystat()`.
* Writing code in a top-down way is like writing an
outline for an essay and then filling in the details. 
    * The main function is the outline.
    * the sub-functions are the details of each topic.

## Exercise

* Create an R script that first defines `standardize_tibble()` and
then `standardize()`. In `standardize()`, replace
`mean()` by a function `center()` and 
`sd()` by a function `spread()`, where `center()` 
and `spread()` are functions that you write 
to compute the mean and SD using only the `sum()` function.
`center()` and `spread()` should remove missing values
by default.

## R packages

- Use the `library()` command to load packages.
- When we load a package
it is inserted in position 2 of the search list, 
just after `.GlobalEnv`.

\footnotesize

```{r}
# install.packages("hapassoc")
library(hapassoc) 
search()
```

## Detaching packages

- Detach a package from the search list with `detach()`

\footnotesize

```{r}
detach("package:hapassoc")
search()
```

## Package namespaces

- Package authors create a list of objects
that will be visible to users when the package
is loaded. This list is called  the package namespace.
- You can access functions in a package's namespace 
without loading the package using
the `::` operator.

\scriptsize

```{r}
set.seed(321)
n<-30; x<-(1:n)/n; y<-rnorm(n,mean=x); ff<-lm(y~x)
car::sigmaHat(ff)
```

\normalsize

- Doing so does not add 
the package to the search list.