---
title: 'Stat 260, Lecture 11, Vectors'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,fig.height=4)
```


## Load packages and datasets

\scriptsize

```{r, message=FALSE, warning=FALSE}
library(tidyverse) 
```

## Reading

* Vectors: Chapter 16 of the printed text or Chapter 20 of the online text.
* Optional: Wickham (2014), Advanced R, Chapter 2,
[http://adv-r.had.co.nz/Data-structures.html/]

## R Data Structures

- There are four main data structures in R: atomic vectors, lists, matrices and data frames.
- Atomic vectors and lists are 1d objects
called vectors, while matrices and data frames are 2d objects
- R has no true scalars; e.g., in `x<-1`, `x` is a vector of length one.
- Use `str()` to see the structure of an object
- Our focus today is on vectors.

## Vector properties

* All vectors have a type and length, which you 
can determine with the `typeof()` and `length()` 
functions, respectively.
* Vectors can have other "attributes".
    * For example, a factor is an integer vector
    with a levels attribute.
    * The text calls such vectors "augmented".

## Types of objects

- Common types we will encounter are "logical", "integer",
"double", "character" and "list".
    * Find the type of an object with `typeof()`.
    
    
\scriptsize

```{r}
x <- 6 # stores as double by default
typeof(x)
y <- 6L # The "L" suffix forces storage as integer  
typeof(y)
```


## Type *versus* Mode

- In addition to the type of an object, there is its
"mode". 
- The mode of an object is generally the same as its type,
but the modes are coarser.
    - Notably, integer and double types are both of mode
    "numeric".

\scriptsize

```{r}
mode(x)
mode(y)
```

## More on numeric variables

* Note that `double`s are floating point
(finite-precision, base-2, with floating decimal place)
approximations of real numbers. 

\scriptsize

```{r}
sqrt(2)^2 - 2
```

\normalsize

* `double`s inlude `NaN`, `Inf`, and `-Inf` for 
division by zero:

\scriptsize

```{r}
c(-1,0,1)/0
```

\normalsize

* What does `NA/0` return? Why does this make sense?

## Creating Vectors
\small

- Vectors can be either atomic or list
    - The elements of an atomic vector must be the **same** type.
    - Lists can be comprised of **multiple** data types
- Empty vectors can be created by the `vector()` function:

\scriptsize

```{r}
# help("vector")
avec <- vector(mode="numeric",length=4)
lvec <- vector(mode="list",length=4)
```

## Creating vectors with `c()` and `list()`

- Data vectors can be created with `c()` or `list()`:

\scriptsize

```{r}
avec <- c(52,200,77)
lvec <- list(52,200,77,c("grey","thin"))
```

## Combining vectors

- Use `c()` to combine vectors

\scriptsize

```{r}
c(avec,c(100,101))
c(lvec,TRUE)
```

## Examples of vector type and length

\scriptsize

```{r}
typeof(avec)
length(avec)
str(avec)
typeof(lvec)
length(lvec)
```

## Named vectors

* Vector elements can have names.
* Names can be assigned after the vector has
been created, or in the process of creating the
vector.

\scriptsize

```{r}
names(lvec) = c("age","weight","height","hair")
str(lvec)
lvec <- list(age=52,weight=200,height=77,hair=c("grey","thin"))
```


## `NULL`

* The absence of a vector is indicated by `NULL`.
* `NULL` is its own type, and is of length 0.

\scriptsize

```{r}
typeof(NULL)
length(NULL)
```

## Exercise

*  Write
a function `append1()` that takes an argument `n`.
The function body should (i) initialize
an object `x` to `NULL`, (ii) loop from i in 1 to `n`
and at each iteration use `c(x,i)` to extend 
`x` by one element, and (ii) return `x`. 
Use the `system.time()` function to time
`append1()`. In particular, compare the following:

```{r,eval=FALSE}
system.time({x <- append1(10000)}) 
system.time({x <- 1:10000})
```



## Subsetting vectors

- Subset with `[` or by name.
- Index values indicate the subset.
- Negative values drop elements.
- Subsetting with a logical vector keeps all
elements where there is a `TRUE` in the logical.:

\scriptsize

```{r}
lvec[c(1,3)] 
#lvec[c("age","height")]
#lvec[-2]
#lvec[c(TRUE,FALSE,TRUE)]
```

## Extracting vector elements

- Extract individual elements with `[[`, or `$` for named objects:

\scriptsize

```{r}
avec[[2]]
lvec[[4]]
lvec$hair
```

* **Exercise** 
 How would you extract 200 from `lvec`?
How would you extract the sub-list containing
weight and height data from `lvec`?
How would you extract `grey` from `lvec`?

## Subsetting and assignment

- You can combine subsetting and assignment to change 
the value of vectors

\scriptsize

```{r}
avec
avec[1:2] <- c(53,215)
avec
```

## Assignment and lists

* To assign to a vector element, use `[[` rather than `[`.
    - This is particularly important with assignments to 
    lists.
    - Assignment with `[` requires that the replacement 
    element be of length 1; `[[` 
    does not have this restriction
    
\scriptsize

```{r}
lvec[3:4] <- c("Hi","there")
lvec[3:4]
```

##

\scriptsize

```{r}
lvec[4] <- c("All","of","this")
lvec[4] # Only used first element of replacement vector
lvec[[4]] <- c("All","of","this")
lvec[3:4]
```


## Coercion: atomic vectors to lists

- Atomic vectors can be coerced to lists with `as.list()`:

\scriptsize

```{r}
avec = c(age=52,weight=200,height=77)
avec
as.list(avec)
```

* **Exercise** The function `as.vector()` 
coerces objects to vectors. Why doesn't  `as.vector(lvec)` 
appear to do anything?



## Coercion: lists to atomic vectors

- Lists can be "flattened" into atomic vectors with `unlist()`:

\scriptsize

```{r}
unlist(lvec)
```

\small

- Notice how the numeric values are coerced to the more flexible 
character type. 
- The order of flexibility, from least to most, is logical,
integer, numeric, character.



## Test functions

* Function outputs may depend on the type of an input object.
* The test functions `as.*`, or their tidyverse equivalents
`as_*` can be used to test object type.
* Useful functions are `is_logical()`, `is_numeric()`,
`is_character()`, `is_list()` and `is_vector()`.


## Recycling

* Arithmetic between a longer and shorter object 
leads to recyling of the shorter object.

\scriptsize

```{r,warning=TRUE}
x <- rep(100,10)
y <- 1:3
x + y
```

\normalsize

* This is a handy way to add a scalar to a vector,
but is dangerous for most anything else.


## Generic functions

* Generic functions behave differently depending 
on the class of input. 
* One of the most important generic functions is 
`print()`.

\scriptsize

```{r}
print
```

\normalsize

* `UseMethod("print")` means that this is a 
generic function that will call different functions
(methods) for objects of different classes.


## Methods for `print()`

* There are many! Here we just print the first 10.
* In addition, there is a default that 
prints any object without a defined method.

\scriptsize


```{r}
methods("print")[1:10]
```

## Seeing methods with `getS3method()`

\scriptsize

```{r}
getS3method("print","default")
```


## Defining your own class

- You can create your own `class` for an object 
and define methods for it.

\scriptsize

```{r}
class(lvec) <- "prof" # print(lvec)
print.prof <- function(p){
  cat("The prof is",p$age,"years old, and weighs",p$weight,"pounds\n")
}
print(lvec)
```

## Exercise

* Create a list of information on this class. 
The list should have named elements to hold the
following information:

\begin{tabular}{l|l}
class day & Tuesday \\
class start & 12:30pm \\
class end & 2:20pm \\
final exam & 	2019/12/07 12:00pm \\
text book & R for Data Science
\end{tabular}

* Use dates or date-times for the times and date-times
in the above. Assign class `SFUcourse` to the
list. Write a function `diff.SFUcourse()` 
that takes an object of class `SFUcourse` as
input and returns the duration of the lecture.
(Here "duration" is as discussed in lecture 9
on dates and times.)

## Augmented vectors

* Vector attributes and classes can be used 
to make useful data structures out of vectors.
* Examples include factors, dates, date-times
and data frames/tibbles.
* For example, a factor is an integer vector
with a `levels` attribute that maps the 
integer values to the factor levels.

\scriptsize

```{r}
ff <- factor(c("a","b","c"))
typeof(ff)
attributes(ff)
```

\normalsize

* See the text for  a description  of
dates and date-times as augmented vectors.

## Data frames and tibbles

* A tibble is an "improved" data frame. 
* Data frames and tibbles are implemented 
as lists with attributed `names`, `row.names`.
* All elements of a tibble or data frame 
must be the same length.

\scriptsize

```{r}
x <- tibble(a=1:3,b=6:8)
attributes(x)
```

\normalsize

* The `tbl_df` and `tbl` aspects of the class are
specific to tibbles. Many methods such as print 
are different for tibbles than data frames, 
but any method not defined is inherited from 
the data frame class.

